\documentclass{report}

\input{~/latex/template/preamble.tex}
\input{~/latex/template/macros.tex}

\title{\Huge{Maps and Unordered Maps}}

\author{\huge{Matt Warner}}
\date{\huge{}}
\pagestyle{fancy}
\fancyhf{}
\rhead{}
\lhead{\leftmark}
\cfoot{\thepage}
% \usepackage[default]{sourcecodepro} \usepackage[T1]{fontenc}

\pgfpagesdeclarelayout{boxed}
{
  \edef\pgfpageoptionborder{0pt}
}
{
  \pgfpagesphysicalpageoptions
  {%
    logical pages=1,%
  }
  \pgfpageslogicalpageoptions{1}
  {
    border code=\pgfsetlinewidth{1.5pt}\pgfstroke,%
    border shrink=\pgfpageoptionborder,%
    resized width=.95\pgfphysicalwidth,%
    resized height=.95\pgfphysicalheight,%
    center=\pgfpoint{.5\pgfphysicalwidth}{.5\pgfphysicalheight}%
  }%
}

\pgfpagesuselayout{boxed}

\begin{document}
  \maketitle
  \section{Maps} 
  Maps can be used by adding the map header file to your program:
  \begin{minted}{c++}
  #include <map>
  \end{minted}
\bigbreak \noindent
Maps are \textbf{associative containers} that store elements in a mapped fashion. Each element has a key value and a mapped value. No two mapped values can have the same key value.
\begin{mdframed}
 Some basic functions associate with std::map are: 
 \bigbreak \noindent
 \begin{itemize}
   \item begin() - Returns an iterator to the first element in the map. 
   \item end() - Returns an iterator to the theoretical element that follows the last element in the map
   \item size() - Returns the number of elements in the map.
   \item max\_size() - Returns the maximum number of elements that the map can hold
   \item empty() - Returns whether the map is empty.
   \item pair insert(keyvalue, mapvalue)  - Adds a new element to the map.
   \item erase(iterator position) - Removes the elements at the position pointed by the iterator
   \item erase(const g) - Removes the key-value 'g' from the map.
   \item clear() - Removes all the elements from the map.
 \end{itemize}
\end{mdframed}
\bigbreak \noindent
\subsection*{Examples of std::map}
The following examples shows how to perform basic operations on map containers
\bigbreak \noindent
\textbf{Example 1: using .begin() and .end()}
\begin{minted}{c++}
#include <iostream>
#include <map>
#include <string>
using namespace std;

int main()
{
  map<string, int> mp;    // Creates a map of strings to integers

  // Insert some values into the map
  mp["one"] = 1;
  mp["two"] = 2;
  mp["three"] = 3;
  
  // Get an iterator pointing to the first element in the map
  map<string, int>::iterator it = mp.begin();

  // Iterate through the map and print the elements
  while (it != mp.end())
  {
    cout << "Key: " << it->first
         << ", Value: " << it->second endl;
    it++;
  }
  return 0;
}
\end{minted}
 \noindent
\textbf{Output:}
\begin{mdframed}
 Key: one, Value: 1 
 \bigbreak \noindent
 Key: three, Value: 3 
 \bigbreak \noindent
 Key: two, Value: 2 
\end{mdframed}
\nt{
  Maps are implemented as a balanced binary tree, and it automatically sorts its elements based on the key.
  \bigbreak \noindent
  The sorting is done in ascending order according to the key's value.
}
\textbf{Example 2: Using size() function}
\begin{minted}{c++}
int main(}
{
  map<string,int> map;

  map["one"] = 1;
  map["two"] = 2;
  map["three"] = 3;

  cout << "Size of map: " << map.size() << endl;

  return 0;
}
\end{minted}
\textbf{Output:}
\begin{mdframed}
  \vspace{.5mm}
  Size of map: 3  
\end{mdframed}
  \vspace{.5mm}
  \textbf{Example 3: inserting elements }
  \begin{minted}{c++}
  std::map<int,int> mp;

  mp.insert(pair<int,int>(1,40));  // First method for inserting elements

  mp[1] = 3;     // Second method
  \end{minted}

\end{document}
