\documentclass{report}

\input{~/latex/template/preamble.tex}
\input{~/latex/template/macros.tex}

\title{\Huge\textbf{Group Project Individual Survey}}

\author{
    \Large Matt Warner
}

\date{\huge{}}
\pagestyle{fancy}
\fancyhf{}
\rhead{}
\lhead{\leftmark}
\cfoot{\thepage}
% \usepackage[default]{sourcecodepro} \usepackage[T1]{fontenc}

\pgfpagesdeclarelayout{boxed}
{
  \edef\pgfpageoptionborder{0pt}
  \pgfpagesphysicalpageoptions
  {%
    logical pages=1,%
  }
  \pgfpageslogicalpageoptions{1}
  {
    border code=\pgfsetlinewidth{1.5pt}\pgfstroke,%
    border shrink=\pgfpageoptionborder,%
    resized width=.95\pgfphysicalwidth,%
    resized height=.95\pgfphysicalheight,%
    center=\pgfpoint{.5\pgfphysicalwidth}{.5\pgfphysicalheight}%
  }%
}

\pgfpagesuselayout{boxed}

\begin{document}
    \maketitle
\section{Contributions to the project}
\begin{itemize}
    \item CSS
    \item Contributed to many of the user interface features.
    \item Added the queue sort + user interface table sort functionality.
    \item Helped with the ERD and relational schema design.
\end{itemize}
\section{Things I learned from the project}
\begin{itemize}
    \item how to create and structure dynamic web pages.
    \item Learned about php syntax and some php built in functions.
    \item Learned how to use php sessions to save user input across multiple refreshes.
    \item Also learned a lot of CSS.
\end{itemize}
\section{Tools used for collaboration}
\begin{itemize}
\item Discord
\item Google drive
\end{itemize}
\section{Team rating}
\begin{enumerate}
    \item Sang Pham \ - \ 10/10
        \begin{itemize}[label=$\circ$]
            \item Main contributor of the DJ interface, also put a lot of work into the user interface and er diagram.
        \end{itemize}
    \item  James Bolla \ - \ 10/10 
        \begin{itemize}[label=$\circ$]
        \item Contributed to the conceptual modeling along with functionality for the user interface. Also contributed to the DDL and DML scripts.
        \end{itemize}
    \item Emanuel Slaughter \ - \ 10/10 
        \begin{itemize}[label=$\circ$]
            \item Contributed to the php code for the user interface and helped design the ER diagram and relational schema.
        \end{itemize}
    \item Ty Danan  \ - \ 10/10
        \begin{itemize}[label=$\circ$]
        \item Contributed to the ER diagram and helped write the DDL and DML scripts, also helped with the user interface.
        \end{itemize}
\end{enumerate}
\end{document}

